Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-18 doc/invt.tex
invt.tex: editing results

The PMF was calculated from $r = 0$ to half of the box size
(also the cutoff distance of the potential)
on a grid of size $\delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
The reduced temperature was $T = 3$.
%
We tested the single-bin, Gaussian,
and homogeneous bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under a constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The
  seemingly small updating magnitude
  corresponds to a larger value of $\ln f = n \, a_0$
  in the WL algorithm.
  The $\ln f$ was $0.05$ for $\varrho = 0.1$,
  or $0.025$ for $\varrho = 0.8$.}
followed by a production phase of another $10^7$ steps
with updating magnitude given by
either the inverse-time or optimal schedule.
%
The deviation of the bias potential was calculated from
Eqs. \eqref{eq:v_def} and \eqref{eq:x_def},
and decomposed to eigenmodes by
Eqs. \eqref{eq:gft_def} and \eqref{eq:wband_eigenvector_refl}.
%
The error was computed from Eq. \eqref{eq:error_def}.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed
$\Gamma_k$'s and $\epsilon_k$'s,
although ideally they could be estimated from the data
collected from the equilibration phase in individual runs.

@@

To illustrate the mode dependence of the error,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index, $k$,
for the $\varrho = 0.1$ case
with width $\sigma = 0.2$.
%
For modes $k=10$ to $25$,
the final errors from the Gaussian updating scheme
were greater than those from the single-bin updating scheme,
although the initial values were less.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to
reduce the updating magnitude too quickly,
leaving a hump on the error profile
for these middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
As a result, the optimal schedule usually produced
less error than the inverse-time schedule,
as shown in Table \ref{tab:lj_error}.
%

One way to overcome the systematic bias
in the Gaussian updating scheme
is to use the histogram-based correction formula,
Eq. \eqref{eq:uhatav}.
%
As shown in Fig. \ref{fig:lj_xerr},
even for the simulation using the non-optimal $1/t$ schedule,
the correction was able to reproduce
the flatter profile seen in the single-bin case.
%
However, this correction also introduced
some local noise in the large $k$ modes,
and it would be best used as a post-simulation
analysis tool for simulations
with detectable residual bias.
%
For example, for the simulation in the $\varrho = 0.8$ case
with $\sigma = 0.5$ and the inverse-time schedule,
the correction was able to lower
the error from $9.8\times10^{-3}$
(Table \ref{tab:lj_error}) to $3.3\times10^{-3}$.

@@

For the Gaussian updating scheme,
there was an optimal width for maximum effectiveness.
%
The optimal width was $\sigma = 0.37$ for $\varrho = 0.1$
or $0.07$ for $\varrho = 0.8$,
and it also depended on the simulation length
and other parameters.
%
Although widening the Gaussian can
effectively suppress random error
from local histogram fluctuation,
it also increased the risk of including
more systematic bias in middle-range modes,
which can be difficult to remove
even with the optimal schedule.
%
This is because the rate of error reduction
is proportional to the eigenvalue, $\lambda_k$
[cf. Eq. \eqref{eq:vt_diffeq_mbin}],
which decreases rapidly with the mode index $k$,
cf. Eq. \eqref{eq:lambda_Gaussian}.

\begin{figure}[h]\centering
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/lj_xerr.pdf}
  }
  \caption{
    \label{fig:lj_xerr}
    Error components from FDS simulations
    using the Gaussian updating scheme
    of width $\sigma = 0.2$
    on the LJ system at density $\varrho = 0.1$.
    %
    The optimal schedule (opt.) was more effective
    in reducing the errors of middle-range eigenmodes
    than the inverse-time schedule ($1/t$).
    The ``corrected'' values were obtained from
    Eq. \eqref{eq:uhatt}.
    The lines are a guide to the eyes.
  }
\end{figure}

The Gaussian updating scheme produced less error than
the single-bin scheme in the $\varrho = 0.1$ case,
but not in the $\varrho = 0.8$ case.
%
The effectiveness of the Gaussian updating scheme
in the low-density case
was probably because the sampling along
the collective variable was free from major barriers
and thus the histogram fluctuation largely
came from local noise.
%
%The optimal schedule also appeared to be more effective
%in reducing the error than the inverse-time schedule.
%%
In this case, we could further reduced the error
by using the bandpass updating scheme
(with the mode cutoff $K = 20$).
%
On the other hand, in the high-density case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range fluctuation,
as $\Gamma_1 \approx \Gamma$,
and the reduction of local noise
had little effect on the total error.

But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in a much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the predicted schedule
would be unusable because the resulting deviation in the bias potential
was no longer small,
and one may have to either decrease the width
or to elongate the preliminary run.

The initial magnitude was, however, around
the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.
%
Since we have chosen a long equilibration period
for the preliminary run under the constant updating magnitude, $a_0$,

>>>>

The PMF was calculated from $r = 0$ to half of the box size
(also the cutoff distance of the potential)
on a grid of size $\delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
The reduced temperature was $T = 3$.
%
We tested the single-bin, Gaussian,
and homogeneous bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under a constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The
  seemingly small updating magnitude
  corresponds to a larger value of $\ln f = n \, a_0$
  in the WL algorithm.
  The $\ln f$ was $0.05$ for $\varrho = 0.1$,
  or $0.025$ for $\varrho = 0.8$.}
followed by a production phase of
another $10^7$ steps (unless indicated otherwise)
with updating magnitude given by
either the inverse-time or optimal schedule.
%
The deviation of the bias potential was calculated from
Eqs. \eqref{eq:v_def} and \eqref{eq:x_def},
and the error from Eq. \eqref{eq:error_def}.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed
$\Gamma_k$'s and $\epsilon_k$'s,
although ideally they could be estimated from the data
collected from the equilibration phase in individual runs.

@@

@@

For the same case,
we further decomposed the error (after $10^7$ production steps)
into the eigenmodes
by Eqs. \eqref{eq:gft_def} and \eqref{eq:wband_eigenvector_refl},
and ploted the error components, $\langle \tilde v_k^2 \rangle$,
in Fig. \ref{fig:lj_xerr}.
%
We found that the inverse-time schedule tended to
reduce the updating magnitude too quickly,
leaving a hump on the error profile
for the $k = 10$ to $25$ modes.
%
By a delayed reduction,
the optimal schedule more effectively reduced
the error in these modes,
and produced a flatter profile for the final error.

\begin{figure}[h]\centering
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/lj_xerr.pdf}
  }
  \caption{
    \label{fig:lj_xerr}
    Error components from FDS simulations
    using the Gaussian updating scheme
    of width $\sigma = 0.2$
    on the LJ system at density $\varrho = 0.1$
    after $10^7$ production steps.
    %
    The optimal schedule (opt.) was more effective
    in reducing the errors of middle-range eigenmodes
    than the inverse-time schedule ($1/t$).
    The ``corrected'' values were obtained from
    Eq. \eqref{eq:uhatt}.
    The lines are a guide to the eyes.
  }
\end{figure}

Next, we will discuss the applicability of
the Gaussian updating scheme.
%
As shown in Table \ref{tab:lj_error},
the Gaussian updating schemes,
or more generally the bandpass updating schemes,
produced less error than the single-bin scheme 
only in the low-density $\varrho = 0.1$ case,
but not in the high-density $\varrho = 0.8$ case.
%
We believe that the difference lies in the sampling difficulty.
%
In the low-density case,
the sampling along the collective variable
was free from major barriers,
and thus the histogram fluctuation largely
was largely local in nature.
%
In the high-density case, however,
the error was dominated by the first few eigenmodes
that correspond to the long-range fluctuations
(as $\Lambda_1 \approx \Lambda$),
and the reduction of local noise by multiple-bin schemes
had little effect on the total error.
%
When the local noise are dominant,
we found that the bandpass updating scheme
(with the mode cutoff $K = 20$)
was more effective than the Gaussian updating schemes
as the expected.

It is worth mentioning that the width $\sigma$
of a Gaussian updating scheme
can critially affect its effectiveness.
%
For example, the prediction showed that
there was a optimal width
around $\sigma = 0.37$ for $\varrho = 0.1$
or $0.07$ for $\varrho = 0.8$,
assuming the optimal schedule,
and the optimal width also depended on the simulation length
and other parameters.
%
We recall that a wider Gaussian means
more middle-range eigemodes
would have diminishing eigenvalues
by Eq. \eqref{eq:lambda_Gaussian}.
%
Thus, although widening the Gaussian can
effectively suppress random error
from local histogram fluctuation,
it also increased the risk of including 
more systematic bias in middle-range modes,
which can be difficult to remove
as the rate of error reduction
is also proportional to the eigenvalue, $\lambda_k$,
by Eq. \eqref{eq:vt_diffeq_mbin}.

An overly large width $\sigma$ can have a dramatic
impact on the initial updating magnitude
under the optimal schedule.
%
The initial updating magnitude, $\alpha(0)$,
is usually around
the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0},
for a sufficiently long preliminary run
under the constant updating magnitude, $a_0$,
as shown in Fig. \ref{fig:lj_alpha}.
%
But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in a much larger $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the predicted optimal schedule
becomes invalid the resulting deviation in the bias potential
is no longer small.

In practice, both the width and schedule
of the Gaussian updating scheme can be sub-optimal.
%
We can use the histogram-based correction formula,
Eq. \eqref{eq:uhatav}, after simulation
to remove the remaining bias in PMF.
%
In Fig. \ref{fig:lj_xerr},
we show a case where the correction produced
the flatter profile seen in the single-bin case,
for a simulation with a sub-optimal schedule.
%
However, this correction will also introduce
some local noise in the large $k$ modes,
although these noise are largely harmless
in a post-simulation analysis, since
the bias potential no longer need to be differentiated
for force calculation.
%
%For example, for the simulation in the $\varrho = 0.8$ case
%with $\sigma = 0.5$ and the inverse-time schedule,
%the correction was able to lower
%the error from $9.8\times10^{-3}$ to $3.3\times10^{-3}$.
